Reassessing the Role of Theory and Machine Learning in Forecasting Civil Conflict

### 2020-09-22 paper.Rmd
finished for the day, this time it's true

We believe that B&S intend that the escalation model's predictions for the test set are weighted by the PITF model predictions for the test set. However, they actually weight the \textbf{test} set predictions using the PITF model predictions for the \textbf{training} set. This appears to be an easily corrected coding error on the part of B&S.

@@

Furthermore, if what we mean with "theory" is such that it suggests a large range of variables that could be related to an outcome, what do we really gain from having this theory in our predictive model? We can construct a model designed to predict well while avoiding overfitting, and then go back and try to couch it in theoretical language---which, given the malleability of extant theory, we probably could in many cases---but are we then better off? Certainly that's not what B&S are trying to suggest since this would be a problematic way to build a theory. But when theory is extensive or malleable so that it suggests a broad range of variables and specifications, and thus a broad range of specifications could post fact be tied to a theory, it is hard to demonstrate that this theory is in fact not the result of predictive modeling's equivalent to p-hacking, namely developing a model specification and any associated "theory" with any eye towards out-of-sample predictive accuracy [cf. mahmood colarsei paper on how to do this properly though]. Incidentally, this is also a recipe for creating overfit models whose accuracy does not generalize, and whose underlying theories are by extension not externally valid [zones farris paper].

Conversely, not all factors that are "theoretically" important or suggested by the literature are also important for prediction, and some factors that turn out to be good predictors do not originate from prior theory. For example, both @wgb:2007 and @hill:jones:2014 examine a variety of variables that are considered "theoretically" important for explaining civil war onset and state repression, respectively, and find that there is substantial variation in how useful they are for for actually predicting those outcomes. And it is entirely possible that a focus on building accurate predictive models could uncover factors overlooked in the literature, e.g. the strong association between infant mortality and a broad variety of conflict outcomes like political instability [@goldstone:etal:2010], irregular leadership changes [@beger:etal:2017], and coups.^[Andreas Beger and Michael D. Ward, 2020, “Coup forecasts for 2020”, https://www.predictiveheuristics.com/forecasts]

@@

Aside from selection of the variables that go into a model, another way in which we could contrast theoretical from atheoretical prediction models is in terms of their flexibility in regard to functional form. Theory should not only tell us what factors matter, but also how they relate to each other. We think that few would be strongly committed to the idea that the linear additive equations that dominate published regression analyses, maybe with the occasional polynomial or multiplicative interaction terms, are *the* correct functional specification. Is the danger posed by low-level violence the same when tensions are high, with high demands and high non-violent repression, as when tensions are low? Are those relationships linear in the log-odds? Some machine learning models, random forests included, are flexible in this regard and to some extent capable of learning non-linear and interactive relationships. B&S implicitly concede on this point since they use random forests for both their Escalation model and the atheoretical alternatives. But in the absence of strong a priori theoretical expectations about functional form, why not use machine learning tools to try to uncover any such non-linear relationships in need of explanation, and use this to improve our theory?

@@

Additional details in issue #3 in the GitHub repo.

>>>>

We believe that B&S intend that the escalation model's predictions for the test set are weighted by the PITF model predictions for the test set. However, they actually weight the \textbf{test} set predictions using the PITF model predictions for the \textbf{training} set.

As a result: (1) in the test set, all countries with names that alphabetically start after "Tongo" are dropped from the data, including Turkey, Ukraine, Yemen, etc.; and (2) the cases whose PITF training data predictions are combined with Escalation model test predictions in the weighting are nonsensical. For example, the "Weighted by PITF" model predictions for Syria from 2008 to 2015 are the Escalation test predictions for that time period weighted, respectively, by the 2002 to 2007 PITF predictions for South Africa and the 2001 to 2002 PITF predictions for Zambia.

This appears to be an easily corrected coding error on the part of B&S.

@@

Furthermore, if what we mean with "theory" is such that it suggests a large range of variables that could be related to an outcome, what do we really gain from having this theory in our predictive model? We can construct a model designed to predict well while avoiding overfitting, and then go back and try to couch it in theoretical language---which, given the malleability of extant theory, we probably could in many cases---but are we then better off? Certainly that's not what B&S are trying to suggest since this would be a problematic way to "evaluate" a theory. But when theory is extensive or malleable so that it suggests a broad range of variables and specifications, and thus a broad range of specifications could post fact be tied to a theory, it is hard to demonstrate that this theory is in fact not the result of predictive modeling's equivalent to p-hacking, namely developing a model specification and associated "theory" with any eye towards out-of-sample predictive accuracy [cf. @colaresi:mahmood:2017 for how to avoid this problem while leveraging out-of-sample predictions]. Incidentally, this is also a recipe for creating overfit models whose accuracy does not generalize, and whose underlying theories may by extension not be externally valid [for a more detailed treatment of this issue, see @fariss:jones:2018].

Conversely, not all factors that are "theoretically" important or suggested by the literature are also important for prediction, and some factors that turn out to be good predictors do not originate from prior theory. For example, both @ward:etal:2010 and @hill:jones:2014 examine a variety of variables that are considered "theoretically" important for explaining civil war onset and state repression, respectively, and find that there is substantial variation in how useful they are for for actually predicting those outcomes. And it is entirely possible that a focus on building accurate predictive models could uncover factors overlooked in the literature, e.g. the strong association between infant mortality and a broad variety of conflict outcomes like political instability [@goldstone:etal:2010], irregular leadership changes [@beger:etal:2017], and coups.^[Andreas Beger and Michael D. Ward, 2020, “Coup forecasts for 2020”, https://www.predictiveheuristics.com/forecasts]

@@

Aside from selection of the variables that go into a model, another way in which we could contrast theoretical from atheoretical prediction models is in terms of their flexibility in regard to functional form. Theory should not only tell us what factors matter, but also how they relate to each other. We think that few would be strongly committed to the idea that the linear additive equations that dominate published regression analyses, maybe with the occasional polynomial or multiplicative interaction terms, are *the* correct functional specification. Is the danger posed by low-level violence the same when tensions are high, with high demands and high non-violent repression, as when tensions are low? Are those relationships linear in the log-odds? Some machine learning models, random forests included, are flexible in this regard and to some extent capable of learning non-linear and interactive relationships. B&S implicitly concede on this point since they use random forests for both their Escalation model and the atheoretical alternatives. But in the absence of strong a priori theoretical expectations about functional form, why not use machine learning tools to try to uncover any such non-linear relationships in need of explanation, and use this to improve our theory [see also @fariss:jones:2018]?

@@

Uncovering which cases in the PITF training predictions are matched to the cases that the Escalation model test set predictions correspond to requires tracing back several more data objects. We have documented details in issue #3 in the GitHub repo.
